\thispagestyle{empty}
\chapter*{Acknowledgements}
\label{acknowledgements}
\vspace{-1em}

\noindent
First and foremost, I am grateful to Florinda De Simini for
encouraging me to apply for a position in her 
\textsc{\hbox{shivadharma} project} (ERC\linebreak no.\thinspace 803624),
for generously sharing all the relevant manuscript materials with me, 
for providing invaluable advice whenever needed, and for 
leading the project in the most friendly and
generous way through happy times as well as 
difficult Covid-affected years. 
While working on the \Vss, I was also affiliated with 
another ERC project, the \textsc{dharma project} (ERC no.\thinspace 809994).
I am grateful to all my colleagues involved in that endeavour,
including Arlo Griffith, Emmanuel Francis, Annette Schmiedchen, 
Astrid Zotter, and Dániel Balogh.

As always, I must express my gratitude to my former supervisor,
Alexis Sanderson, and to Dominic Goodall and 
Harunaga Isaacson, for initiating me into Sanskrit philology and
the study of Śaivism.

My colleagues and friends working in Naples or visiting for
shorter periods helped me on a daily basis, during our regular reading sessions and in countless
other ways. I am thankful to them (in alphabetical order):
Alessandro Battistini, 
Michael Bluett, 
Lucas den Boer, 
Martina Dello\linebreak Buono, 
Giulia Buriola, 
Daniela Cappello,
Csaba Dezső, 
Marco\linebreak \mbox{Franceschini}, 
Torsten Gerloff, 
Kengo Harimoto, 
Gergely Hidas,
Nirajan Kafle, 
Chiara Livio, 
Timothy Lubin, 
Nina Mirnig, 
Andrew Ollett,\linebreak 
Dorotea Operato, 
Martin Orwin, 
Luca Piscopo, 
Alexander von Rospatt, 
S.\thinspace A.\thinspace S. Sarma,
R.\thinspace Sathyanarayanan, 
Francesco Sferra, 
%Florinda De Simini,
Kenji Takahashi,\linebreak 
Judit Törzsök,
Margherita Trento, 
and others.
% students in Naples?
I am grateful to Daniela Cappello, Marco Franceschini, 
and Sushmita Das for their great efforts in acquiring manu\-scripts in Calcutta.

During my visit to the National Archives in
Kathmandu, the staff were as helpful and professional as ever.
I wish to express my thanks to Jyoti\linebreak Neupane, Manita Neupane, Saubhagya Pradhananga, Rubin Shrestha,\linebreak
Sahan Ranjitkar, and all %sribin7501@gmail.com
other members of the team.

I thank my host in Capodimonte, Michele Costagliola, for his generosity. 
I am infinitely grateful to my family for always supporting me unwaveringly.

\vspace{2cm}
\noindent

\CHECK\ REVISE!!!

{\footnotesize
The present publication is a result of the project \textsc{dharma} `The
Domestication of ``Hindu'' Asceticism and the Religious Making of South and Southeast Asia'.  This project has received funding from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation programme (grant agreement no.\thinspace 809994).  This book reflects the views of the author only.  The funding body is not responsible for any use that
may be made of the information contained therein.}

%CHECK what to add for the Śivadharma project?

\vfill
\pagebreak
